---
title: "Project 3: Preference and Choice / Market Basket Analysis"
author: "Marjorie Blanco"
subtitle: DA 420
output:
  pdf_document:
    fig_caption: yes
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(support.CEs)  # package for survey construction 
library(arules)       # association rules
library(arulesViz)     # data visualization of association rules
library(RColorBrewer)  # color palettes for plots
library(knitr)
source("Chapter_3/R_utility_program_1.R")
library(kableExtra)
library(kableExtra)
print.digits <- 2    # set number of digits on print and spine chart
```


```{r}
plotRule <- function(rules, product, filename) {
  # select rules with product in consequent (right-hand-side) item subsets
  product.rules <- subset(rules, subset = rhs %pin% product)
  
  # sort by lift and identify the top 10 rules
  top.product.rules <- head(sort(product.rules, decreasing = TRUE, by = "lift"), 10)
  inspect(top.product.rules) 
  
#  pdf(file=filename, width = 11, height = 8.5)
  plot(top.product.rules, method="graph", 
       shading = "lift",
       main = cat("Rules for", product, sep = " "))
#  dev.off() 
  return (top.product.rules)
}
```


## Part 1

### Read mobile data

```{r, echo=FALSE}
conjoint.data.frame <- read_csv("data/mobile.csv")
kable(head(conjoint.data.frame))  %>%
  kable_styling(bootstrap_options = "striped")
```

```{r}
# user-defined function for plotting descriptive attribute names 
effect.name.map <- function(effect.name) { 
  if(effect.name=="brand") return("Mobile Service Provider:")
  if(effect.name=="startup") return("Start-up Cost:")
  if(effect.name=="monthly") return("Monthly Cost:")
  if(effect.name=="service") return("Offers 4G Service:")
  if(effect.name=="retail") return("Has Nearby Retail Store:")
  if(effect.name=="apple") return("Sells Apple Products:")
  if(effect.name=="samsung") return("Sells Samsung Products:")
  if(effect.name=="google") return("Sells Google/Nexus Products:")
} 

# set up sum contrasts for effects coding as needed for conjoint analysis
options(contrasts=c("contr.sum", "contr.poly"))

# main effects model specification
main.effects.model <- {ranking ~ brand + startup + monthly + service + 
    retail + apple + samsung + google}

# fit linear regression model using main effects only (no interaction terms)
main.effects.model.fit <- lm(main.effects.model, data=conjoint.data.frame)
print(summary(main.effects.model.fit)) 

# save key list elements of the fitted model as needed for conjoint measures
conjoint.results <- 
  main.effects.model.fit[c("contrasts", "xlevels", "coefficients")]

conjoint.results$attributes <- names(conjoint.results$contrasts)

# compute and store part-worths in the conjoint.results list structure
part.worths <- conjoint.results$xlevels  # list of same structure as xlevels
end.index.for.coefficient <- 1  # intitialize skipping the intercept
part.worth.vector <- NULL # used for accumulation of part worths
for(index.for.attribute in seq(along=conjoint.results$contrasts)) {
  nlevels <- length(unlist(conjoint.results$xlevels[index.for.attribute]))
  begin.index.for.coefficient <- end.index.for.coefficient + 1
  end.index.for.coefficient <- begin.index.for.coefficient + nlevels -2
  last.part.worth <- -sum(conjoint.results$coefficients[
    begin.index.for.coefficient:end.index.for.coefficient])
  part.worths[index.for.attribute] <- 
    list(as.numeric(c(conjoint.results$coefficients[
      begin.index.for.coefficient:end.index.for.coefficient],
      last.part.worth)))
  part.worth.vector <- 
    c(part.worth.vector,unlist(part.worths[index.for.attribute]))    
} 
conjoint.results$part.worths <- part.worths

# compute standardized part-worths
standardize <- function(x) {(x - mean(x)) / sd(x)}
conjoint.results$standardized.part.worths <- 
  lapply(conjoint.results$part.worths,standardize)

# compute and store part-worth ranges for each attribute 
part.worth.ranges <- conjoint.results$contrasts
for(index.for.attribute in seq(along=conjoint.results$contrasts)) 
  part.worth.ranges[index.for.attribute] <- 
  dist(range(conjoint.results$part.worths[index.for.attribute]))
conjoint.results$part.worth.ranges <- part.worth.ranges

sum.part.worth.ranges <- sum(as.numeric(conjoint.results$part.worth.ranges))

# compute and store importance values for each attribute 
attribute.importance <- conjoint.results$contrasts
for(index.for.attribute in seq(along=conjoint.results$contrasts)) 
  attribute.importance[index.for.attribute] <- 
  (dist(range(conjoint.results$part.worths[index.for.attribute]))/
     sum.part.worth.ranges) * 100
conjoint.results$attribute.importance <- attribute.importance

# data frame for ordering attribute names
attribute.name <- names(conjoint.results$contrasts)
attribute.importance <- as.numeric(attribute.importance)
temp.frame <- data.frame(attribute.name,attribute.importance)
conjoint.results$ordered.attributes <- 
  as.character(temp.frame[sort.list(
    temp.frame$attribute.importance,decreasing = TRUE),"attribute.name"])

# respondent internal consistency added to list structure
conjoint.results$internal.consistency <- summary(main.effects.model.fit)$r.squared 

# user-defined function for printing conjoint measures
if (print.digits == 2) 
  pretty.print <- function(x) {sprintf("%1.2f",round(x,digits = 2))} 
if (print.digits == 3) 
  pretty.print <- function(x) {sprintf("%1.3f",round(x,digits = 3))} 

# report conjoint measures to console 
# use pretty.print to provide nicely formated output
for(k in seq(along=conjoint.results$ordered.attributes)) {
  cat("\n","\n")
  cat(conjoint.results$ordered.attributes[k],"Levels: ",
      unlist(conjoint.results$xlevels[conjoint.results$ordered.attributes[k]]))
  
  cat("\n"," Part-Worths:  ")
  cat(pretty.print(unlist(conjoint.results$part.worths
                          [conjoint.results$ordered.attributes[k]])))
  
  cat("\n"," Standardized Part-Worths:  ")
  cat(pretty.print(unlist(conjoint.results$standardized.part.worths
                          [conjoint.results$ordered.attributes[k]])))  
  
  cat("\n"," Attribute Importance:  ")
  cat(pretty.print(unlist(conjoint.results$attribute.importance
                          [conjoint.results$ordered.attributes[k]])))
}
```

```{r fig.width=8.5, fig.height=5, echo=TRUE}
# plotting of spine chart begins here
spine.chart(conjoint.results)
```

```{r}
pdf(file="mobile_spine_plot.pdf", width = 11, height = 8.5)
spine.chart(conjoint.results)
dev.off()
```


## Part 2

```{r, echo=FALSE}
# grocery transactions object from arules package
data(Groceries)  
```

The data set consists of `r nrow(Groceries)` market baskets across `r ncol(Groceries)` generically-labeled grocery items. 


```{r}
groceries <- aggregate(Groceries, itemInfo(Groceries)[["level2"]])  

# select association rules using thresholds for support and confidence 
rules <- apriori(groceries, 
                 parameter = list(support = 0.025, confidence = 0.05))
print(summary(rules))  # yields 344 rules
```

### Vegetables

```{r echo=TRUE}
plotRule(rules, "vegetables", "market_basket_vegetables_rules.pdf")
```

#### Top 3 Rules

- Customers who buy beef and dairy produce are 2.2 times as likely to buy vegetables as customers from the entire data set.

- Customers who buy poultry are 2.1 times as likely to buy vegetables as customers from the entire data set.

- Customers who buy dairy produce, fruit, and sausage  are 2.1 times as likely to buy vegetables as customers from the entire data set.

To increase profits, I recomend to promote a dinner kit containing meat (beef, poultry or sausage) with vegetable mix.

### Fruit

```{r}
plotRule(rules, "fruit", "market_basket_fruit_rules.pdf")
```


#### Top 3 Rules

- Customers who buy dairy produce, sausage, and vegetables are 2.0 times as likely to buy fruit as customers from the entire data set.

- Customers who buy dairy produce, non-alc. drinks, and vegetables are 2.0 times as likely to buy fruit as customers from the entire data set.

- Customers who buy bread and backed goods, dairy produce, and vegetables are 1.9 times as likely to buy fruit as customers from the entire data set.

To increase profits, I recomend to promote a fruit, cheese and vegetables platter.


### Meat: Pork

```{r echo=TRUE}
plotRule(rules, "pork", "market_basket_pork_rules.pdf")
```

- Customers who buy vegetables are 1.9 times as likely to buy pork as customers from the entire data set.

To increase profits, I recomend to promote a dinner kit containing meat (pork) with vegetable mix.

### Meat: Beef

```{r echo=TRUE}
plotRule(rules, "beef", "market_basket_beef_rules.pdf")
```

#### Top 3 Rules

- Customers who buy dairy produce and vegetables are 2.1 times as likely to buy beef as customers from the entire data set.

- Customers who buy vegetables are 2.0 times as likely to buy beef as customers from the entire data set.

- Customers who buy bread and backed goods and dairy produce are 1.7 times as likely to buy beef as customers from the entire data set.


To increase profits, I recomend to promote a dinner kit containing meat (beef) with vegetable mix.

```{r echo=TRUE}
plotRule(rules, "poultry", "market_basket_poultry_rules.pdf")
```

- Customers who buy vegetables are 2.1 times as likely to buy poultry as customers from the entire data set.


To increase profits, I recomend to promote a dinner kit containing meat (poultry) with vegetable mix.

### Dairy producer

```{r echo=TRUE}
plotRule(rules, "cheese", "market_basket_cheese_rules.pdf")
```

#### Top 3 Rules

- Customers who buy dairy produce and sausage are 2.28 times as likely to buy cheese as customers from the entire data set.

- Customers who buy bread and backed goods, and sausage are 2.2 times as likely to buy cheese as customers from the entire data set.

- Customers who buy bread and backed goods, and fruit are 2.17 times as likely to buy cheese as customers from the entire data set.

```{r echo=TRUE}
plotRule(rules, "dairy produce", "market_basket_dairy_produce_rules.pdf")
```

#### Top 3 Rules

- Customers who buy bread and backed goods, fruit, and sausage are 1.8 times as likely to buy dairy produce as customers from the entire data set.

- Customers who buy bread and backed goods, fruit, and vegetable are 1.79 times as likely to buy dairy produce as customers from the entire data set.

- Customers who buy fruit, sausage, and vegetables are 1.78 times as likely to buy dairy produce as customers from the entire data set.


### Beer

```{r echo=TRUE}
plotRule(rules, "beer", "market_basket_beer_rules.pdf")
```
In general, association rules with lift values close to 1 will are not interesting nor useful than rules with lift values higher than 1. 

### Recomendation


