---
title: "Project 4: Economic Data Analysis"
author: "Marjorie Blanco"
subtitle: DA 420
output:
  pdf_document:
    fig_caption: yes
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
library(lubridate)
library(dplyr)
library('forecast')
library('tseries')
library(kableExtra)
```


```{r}
# Analysis of Economic Time Series (R)

# economic time series gathered with this program are continually
# updated... so predictive models, forecasts, and data visualizations
# produced by this program may differ from those shown in the book

library(quantmod) # use for gathering and charting economic data
library(lubridate) # date functions
library(latticeExtra) # package used for horizon plot
library(forecast) # functions for time series forecasting 
library(lmtest) # for Granger test of causality
```


```{r}
recessions.df = read.table(textConnection(
"Peak, Trough
1857-06-01, 1858-12-01
1860-10-01, 1861-06-01
1865-04-01, 1867-12-01
1869-06-01, 1870-12-01
1873-10-01, 1879-03-01
1882-03-01, 1885-05-01
1887-03-01, 1888-04-01
1890-07-01, 1891-05-01
1893-01-01, 1894-06-01
1895-12-01, 1897-06-01
1899-06-01, 1900-12-01
1902-09-01, 1904-08-01
1907-05-01, 1908-06-01
1910-01-01, 1912-01-01
1913-01-01, 1914-12-01
1918-08-01, 1919-03-01
1920-01-01, 1921-07-01
1923-05-01, 1924-07-01
1926-10-01, 1927-11-01
1929-08-01, 1933-03-01
1937-05-01, 1938-06-01
1945-02-01, 1945-10-01
1948-11-01, 1949-10-01
1953-07-01, 1954-05-01
1957-08-01, 1958-04-01
1960-04-01, 1961-02-01
1969-12-01, 1970-11-01
1973-11-01, 1975-03-01
1980-01-01, 1980-07-01
1981-07-01, 1982-11-01
1990-07-01, 1991-03-01
2001-03-01, 2001-11-01
2007-12-01, 2009-06-01"), sep=',',
colClasses=c('Date', 'Date'), header=TRUE)
```

## Part 1: 

Use R to develop a Multiple Time Series Plot, as seen in Figure 5.1. Make sure you include the command lines and the output/results. Interpret the graphin details. 

Hint:Look at the Exhibit 5.1 in page 62-63.

### Data

```{r echo=TRUE}
# Economic Data from Federal Reserve Bank of St. Louis (FRED system)
# National Civilian Unemployment Rate (monthly, percentage)
getSymbols("UNRATENSA", src="FRED", return.class = "xts")
# Manufacturers' New Orders: Durable Goods (millions of dollars) 
getSymbols("DGORDER", src="FRED", return.class = "xts")
# University of Michigan Index of Consumer Sentiment (1Q 1966 = 100)
getSymbols("UMCSENT", src="FRED", return.class = "xts")
# New Homes Sold in the US, not seasonally adjusted (monthly, millions)
getSymbols("HSN1FNSA",src="FRED",return.class = "xts")
```

```{r echo=TRUE}
ER <- 100 - UNRATENSA # convert to employment rate
dimnames(ER)[2] <- "ER"
ER.data.frame <- as.data.frame(ER)
ER.data.frame$date <- ymd(rownames(ER.data.frame))

ER.time.series <- ts(ER.data.frame$ER, 
  start = c(year(min(ER.data.frame$date)),month(min(ER.data.frame$date))),
  end = c(year(max(ER.data.frame$date)),month(max(ER.data.frame$date))),
  frequency=12)

recessions.trim <- subset(recessions.df, Peak >= min(ER.data.frame$date) )

ggplot(ER.data.frame) + 
  geom_line(aes(x=date, y=ER)) +
  theme_bw() +
  geom_rect(data=recessions.trim, aes(xmin=Peak, xmax=Trough, ymin=-Inf, ymax=+Inf),
            fill='gray', alpha=0.2) +
  labs(title = "National Civilian Employment Rate (monthly, percentage)", 
       caption = "Source: U.S. Bureau of Labor Statistics\nShaded areas indicate U.S. recessions", 
       x = "", y = "Employment Rate") 

decomp = stl(ER.time.series, s.window="periodic")
plot(decomp)
Acf(ER.time.series)
Pacf(ER.time.series)
```
ACF plot shows a spike at lags 1 through 30. PACF plot shows a spike at lags 1 and 13.  This chart shows that the employement rate are cyclical related to the recessions.  The employement rate shows a upward trend after recession years and downward trend during recession years.  The trend and cyclical nature of employment rate is expected.


```{r echo=TRUE}
DGO <- DGORDER/1000 # convert to billions of dollars
dimnames(DGO)[2] <- "DGO" # use simple name for index
DGO.data.frame <- as.data.frame(DGO)
DGO.data.frame$DGO <- DGO.data.frame$DGO
DGO.data.frame$date <- ymd(rownames(DGO.data.frame))
DGO.time.series <- ts(DGO.data.frame$DGO, 
  start = c(year(min(DGO.data.frame$date)),month(min(DGO.data.frame$date))),
  end = c(year(max(DGO.data.frame$date)),month(max(DGO.data.frame$date))),
  frequency=12)

recessions.trim <- subset(recessions.df, Peak >= min(DGO.data.frame$date) )

ggplot(DGO.data.frame) + 
  geom_line(aes(x=date, y=DGO)) +
  theme_bw() +
  geom_rect(data=recessions.trim, aes(xmin=Peak, xmax=Trough, ymin=-Inf, ymax=+Inf),
            fill='gray', alpha=0.2) +
  labs(title = "Manufacturers' New Orders: Durable Goods (millions of dollars)", 
       caption = "Source: U.S. Bureau of Labor Statistics\nShaded areas indicate U.S. recessions", 
       x = "", y = "Durable Goods (millions of dollars)") 

decomp = stl(DGO.time.series, s.window="periodic")
plot(decomp)
Acf(DGO.time.series)
Pacf(DGO.time.series)
```
ACF plot shows a spike at lags 1 through 30. PACF plot shows a spike at lags 1 and 2.  This chart shows that the manufacturers' new orders are cyclical related to the recessions.  The durable goods shows a upward trend after recession years and downward trend during recession years.  The trend and cyclical nature is expected.

```{r echo=TRUE}
ICS <- UMCSENT # use simple name for xts object
dimnames(ICS)[2] <- "ICS" # use simple name for index
ICS.data.frame <- as.data.frame(ICS)
ICS.data.frame$ICS <- ICS.data.frame$ICS
ICS.data.frame$date <- ymd(rownames(ICS.data.frame))
ICS.data.frame <-  ICS.data.frame[303:nrow(ICS.data.frame),]

ICS.time.series <- ts(ICS.data.frame$ICS, 
  start = c(year(min(ICS.data.frame$date)), month(min(ICS.data.frame$date))),
  end = c(year(max(ICS.data.frame$date)),month(max(ICS.data.frame$date))),
  frequency=12)

recessions.trim <- subset(recessions.df, Peak >= ymd("1978-01-01") )

ggplot(ICS.data.frame) + 
  geom_line(aes(x=date, y=ICS)) +
  theme_bw() +
  geom_rect(data=recessions.trim, aes(xmin=Peak, xmax=Trough, ymin=-Inf, ymax=+Inf),
            fill='gray', alpha=0.2) +
  labs(title = "University of Michigan Index of Consumer Sentiment", 
       caption = "Source: U.S. Bureau of Labor Statistics\nShaded areas indicate U.S. recessions", 
       x = "", y = "Consumer Sentiment Index") 

decomp = stl(ICS.time.series, s.window="periodic")
plot(decomp)
Acf(ICS.time.series)
Pacf(ICS.time.series)
```

ACF plot shows a spike at lags 1 through 30. PACF plot shows a spike at lags 1.  This chart shows that consumer sentiment index is related to the recessions.  The durable goods shows a upward trend after recession years and downward trend during recession years.  The trend and cyclical nature is expected.


```{r echo=TRUE}
NHS <- HSN1FNSA
dimnames(NHS)[2] <- "NHS" # use simple name for index
NHS.data.frame <- as.data.frame(NHS)
NHS.data.frame$NHS <- NHS.data.frame$NHS
NHS.data.frame$date <- ymd(rownames(NHS.data.frame))
NHS.time.series <- ts(NHS.data.frame$NHS, 
  start = c(year(min(NHS.data.frame$date)),month(min(NHS.data.frame$date))),
  end = c(year(max(NHS.data.frame$date)),month(max(NHS.data.frame$date))),
  frequency=12)

recessions.trim <- subset(recessions.df, Peak >= ymd("1978-01-01") )

ggplot(NHS.data.frame) + 
  geom_line(aes(x=date, y=NHS)) +
  theme_bw() +
  geom_rect(data=recessions.trim, aes(xmin=Peak, xmax=Trough, ymin=-Inf, ymax=+Inf),
            fill='gray', alpha=0.2) +
  labs(title = "New Homes Sold in the US, not seasonally adjusted (monthly, millions)", 
       caption = "Source: U.S. Bureau of Labor Statistics\nShaded areas indicate U.S. recessions", 
       x = "", y = "New Homes Sold in the US (millions)") 

decomp = stl(NHS.time.series, s.window="periodic")
plot(decomp)
Acf(NHS.time.series)
Pacf(NHS.time.series)

min_rate1 <- min(NHS.time.series)
max_rate1 <- max(NHS.time.series)

min_date <- NHS.data.frame %>% filter(NHS == min_rate1)
max_date <- NHS.data.frame %>% filter(NHS == max_rate1)
```

ACF plot shows a spike at lags 1, 12 and 14. PACF plot shows a spike at lags 1 and 13.  This chart shows that new homes sold in the US exhibit a seasonal patterns. New homes sold were steadly increasing until  `r max_date` just before the housing market crash.  The lowest home sold in the US was in `r min_date`.

```{r echo=TRUE}
# define multiple time series object
economic.mts <- cbind(ER.time.series, DGO.time.series, ICS.time.series,
  NHS.time.series) 
  dimnames(economic.mts)[[2]] <- c("ER","DGO","ICS","NHS") # keep simple names 
modeling.mts <- na.omit(economic.mts) # keep overlapping time intervals only

# plot multiple time series 
plot(modeling.mts,main="")
```


### Data

The source of this data is the U.S. Bureau of Labor Statistics.  The rate is released monthly as a percent.  The data is seasonally adjusted.

Frequency:  Monthly

```{r echo=TRUE}
# Economic Data from Federal Reserve Bank of St. Louis (FRED system)
# Civilian Labor Force Participation Rate: Women  (monthly, percentage)
getSymbols("LNS11300002", src="FRED", return.class = "xts")
# Civilian Labor Force Participation Rate: Men  (monthly, percentage)
getSymbols("LNS11300001", src="FRED", return.class = "xts")
# Civilian Labor Force Participation Rate: White  (monthly, percentage)
getSymbols("LNS11300003", src="FRED", return.class = "xts")
# Civilian Labor Force Participation Rate: Black  (monthly, percentage)
getSymbols("LNS11300006", src="FRED", return.class = "xts")
# Civilian Labor Force Participation Rate: Hispanic or Latino  (monthly, percentage)
getSymbols("LNS11300009", src="FRED", return.class = "xts")


PR <- LNS11300002 # convert to participation rate
dimnames(PR)[2] <- "PR"
chartSeries(PR, theme="white", name="Participation Rate: Women")
PRWomen.data.frame <- as.data.frame(PR)
PRWomen.data.frame$date <- ymd(rownames(PRWomen.data.frame))

PRWomen.time.series <- ts(PRWomen.data.frame$PR, 
  start = c(year(min(PRWomen.data.frame$date)),month(min(PRWomen.data.frame$date))),
  end = c(year(max(PRWomen.data.frame$date)),month(max(PRWomen.data.frame$date))),
  frequency=12)

PR <- LNS11300001 # convert to participation rate
dimnames(PR)[2] <- "PR"
chartSeries(PR, theme="white", name="Participation Rate: Men")
PRMen.data.frame <- as.data.frame(PR)
PRMen.data.frame$date <- ymd(rownames(PRMen.data.frame))


PRMen.time.series <- ts(PRMen.data.frame$PR, 
  start = c(year(min(PRMen.data.frame$date)),month(min(PRMen.data.frame$date))),
  end = c(year(max(PRMen.data.frame$date)),month(max(PRMen.data.frame$date))),
  frequency=12)

PR <- LNS11300003 # convert to participation rate
dimnames(PR)[2] <- "PR"
chartSeries(PR, theme="white", name="Participation Rate: White")
PRWhite.data.frame <- as.data.frame(PR)
PRWhite.data.frame$date <- ymd(rownames(PRWhite.data.frame))


PRWhite.time.series <- ts(PRWhite.data.frame$PR, 
  start = c(year(min(PRWhite.data.frame$date)),month(min(PRWhite.data.frame$date))),
  end = c(year(max(PRWhite.data.frame$date)),month(max(PRWhite.data.frame$date))),
  frequency=12)

PR <- LNS11300006 # convert to participation rate
dimnames(PR)[2] <- "PR"
chartSeries(PR, theme="white", name="Participation Rate: Black")
PRBlack.data.frame <- as.data.frame(PR)
PRBlack.data.frame$date <- ymd(rownames(PRBlack.data.frame))

PRBlack.time.series <- ts(PRBlack.data.frame$PR, 
  start = c(year(min(PRBlack.data.frame$date)),month(min(PRBlack.data.frame$date))),
  end = c(year(max(PRBlack.data.frame$date)),month(max(PRBlack.data.frame$date))),
  frequency=12)

PR <- LNS11300009 # convert to participation rate
dimnames(PR)[2] <- "PR"
chartSeries(PR, theme="white", name="Participation Rate: Latino")
PRLatino.data.frame <- as.data.frame(PR)
PRLatino.data.frame$date <- ymd(rownames(PRLatino.data.frame))

PRLatino.time.series <- ts(PRLatino.data.frame$PR, 
  start = c(year(min(PRLatino.data.frame$date)),month(min(PRLatino.data.frame$date))),
  end = c(year(max(PRLatino.data.frame$date)),month(max(PRLatino.data.frame$date))),
  frequency=12)
```

### Multiple time series

```{r}
# define multiple time series object
economic.mts <- cbind(PRWhite.time.series, PRBlack.time.series,
  PRLatino.time.series, PRMen.time.series, PRWomen.time.series) 
  dimnames(economic.mts)[[2]] <- c("White","Black","Latino","Men", "Women") # keep simple names 
modeling.mts <- na.omit(economic.mts) # keep overlapping time intervals only

# plot multiple time series 
plot(modeling.mts,main="")
```

The civilian labor force participation rate for Women, White, Black and Latino has been steadily increasing while the rate for Men has been decreasing. The rates for all groups have been decreasing since the last recession.  Data for other races are not available.


```{r}
PR <- LNS11300002 # convert to participation rate
dimnames(PR)[2] <- "PR"
PR.data.frame <- as.data.frame(PR)
PR.data.frame$date <- ymd(rownames(PR.data.frame))

recessions.df = read.table(textConnection(
"Peak, Trough
1857-06-01, 1858-12-01
1860-10-01, 1861-06-01
1865-04-01, 1867-12-01
1869-06-01, 1870-12-01
1873-10-01, 1879-03-01
1882-03-01, 1885-05-01
1887-03-01, 1888-04-01
1890-07-01, 1891-05-01
1893-01-01, 1894-06-01
1895-12-01, 1897-06-01
1899-06-01, 1900-12-01
1902-09-01, 1904-08-01
1907-05-01, 1908-06-01
1910-01-01, 1912-01-01
1913-01-01, 1914-12-01
1918-08-01, 1919-03-01
1920-01-01, 1921-07-01
1923-05-01, 1924-07-01
1926-10-01, 1927-11-01
1929-08-01, 1933-03-01
1937-05-01, 1938-06-01
1945-02-01, 1945-10-01
1948-11-01, 1949-10-01
1953-07-01, 1954-05-01
1957-08-01, 1958-04-01
1960-04-01, 1961-02-01
1969-12-01, 1970-11-01
1973-11-01, 1975-03-01
1980-01-01, 1980-07-01
1981-07-01, 1982-11-01
1990-07-01, 1991-03-01
2001-03-01, 2001-11-01
2007-12-01, 2009-06-01"), sep=',',
colClasses=c('Date', 'Date'), header=TRUE)

recessions.trim <- subset(recessions.df, Peak >= min(PR.data.frame$date) )

recessions.2007 <- PR.data.frame %>% filter(between(date, as.Date("2007-12-01"), as.Date("2009-06-01")))
after.2010 <- PR.data.frame %>% filter(between(date, as.Date("2009-06-01"), as.Date("2018-12-01")))

min_rate1 <- min(recessions.2007$PR)
max_rate1 <- max(recessions.2007$PR)

min_rate2 <- min(after.2010$PR)
max_rate2 <- max(after.2010$PR)
min_date <- after.2010 %>% filter(PR == min_rate2)
```

```{r echo=TRUE}
ggplot(PR.data.frame) + 
  geom_line(aes(x=date, y=PR)) +
  theme_bw() +
  geom_rect(data=recessions.trim, aes(xmin=Peak, xmax=Trough, ymin=-Inf, ymax=+Inf),
            fill='pink', alpha=0.2) +
  labs(title = "Civilian Labor Force Participation Rate: Women ", 
       caption = "Source: U.S. Bureau of Labor Statistics\nShaded areas indicate U.S. recessions", 
       x = "", y = "Participation Rate") 


PR.time.series <- ts(PR.data.frame$PR, 
  start = c(year(min(PR.data.frame$date)),month(min(PR.data.frame$date))),
  end = c(year(max(PR.data.frame$date)),month(max(PR.data.frame$date))),
  frequency=12)
```

### Data decomposition

```{r}
decomp = stl(PR.time.series, s.window="periodic")
plot(decomp)
```

### Check for Stationarity

```{r}
adf.test(PR.time.series, alternative = "stationary")
```

The data is non-stationary; the labor participation rates changes through time. The ADF test does not reject the null hypothesis of non-stationarity, confirming the visual inspection.

### Autocorrelations and Choosing Model Order

```{r}
Acf(PR.time.series)
```

ACF plotshows a spike at lags 1 through 30.

```{r}
Pacf(PR.time.series)
```

PACF plot only shows a spike at lags 1.


### Fitting an ARIMA model

```{r echo=TRUE}
# ARIMA model fit to the partipation rate data
PR.auto.arima.fit <- auto.arima(PR.time.series, d=NA, D=NA, max.p=5, max.q=5,
  max.P=2, max.Q=2, max.order=3, start.p=1, start.q=1,
  start.P=1, start.Q=1, stationary=FALSE, seasonal=FALSE,
  ic=c("aic"), stepwise=TRUE, trace=FALSE,
  approximation=FALSE, xreg=NULL,
  test=c("kpss","adf","pp"), #seasonal.test=c("ocsb","ch"),
  allowdrift=FALSE, lambda=NULL, parallel=FALSE, num.cores=NULL)
print(summary(PR.auto.arima.fit))
# national partipation rate two-year forecast (horizon h = 24 months) 
PR.forecast <- forecast(PR.auto.arima.fit, h=24, level=c(95), 
  fan=FALSE, xreg=NULL, bootstrap=FALSE)
```

AR(1) coefficient p = -1.31 tells us that the next value in the series is taken as a dampened previous value by a factor of -1.32 and depends on previous error lag.

###  Forecast

```{r echo=TRUE}
# plot national partipation rate time series with two-month forecast 
plot(PR.forecast,main="", ylab="Partipation Rate",
  xlab = "Time", las = 1, lwd = 1.5)
```

```{r}
kable(PR.forecast)  %>%
  kable_styling(bootstrap_options = "striped")
```

### Interpretation

The civilian labor force participation rate for Women has been steadily increasing.  The slow decrease/decay in the ACF as the lags increase is due to the upward trend.  During the 2007-2009 recession (FROM 2007-12-01 TO 2009-06-01) the partipation rate was between `r min_rate1` and `r max_rate1`. After the recession the rate has been steadily decreasing. In the last decade the the participation rate was between `r min_rate2` and `r max_rate2` with lowest participation rate on `r month(min_date$date)`-`r year(min_date$date)`.  The 24 month forecast shows the rate to remain steady.
