---
title: 'Project 1: Analytics and Data Science'
author: "Marjorie Blanco"
subtitle: DA 420
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(knitr)
```

# Anscombe Quartet

```{r}
# Read Anscombe data 
anscombe <- read.csv("data/Anscombe.csv")
kable(anscombe)
```

## Descriptive statistics

```{r}
# calculate the mean and standard deviation
kable(anscombe %>% select(-Observation) %>%
    summarise_all(funs( mean(.), sd(.))) %>%
  gather())
```

## Linear Model

```{r}
fit.model <- lm(y1~x1, anscombe)
summary(fit.model)

anscombe$x1_1 <- anscombe$x1 
anscombe$y1_1 <- fit.model$fitted.values -  fit.model$residuals

fit.model <- lm(y2~x2, anscombe)
summary(fit.model)

anscombe$x2_1 <- anscombe$x2 
anscombe$y2_1 <- fit.model$fitted.values -  fit.model$residuals

fit.model <- lm(y3~x3, anscombe)
summary(fit.model)

anscombe$x3_1 <- anscombe$x3 
anscombe$y3_1 <- fit.model$fitted.values -  fit.model$residuals

fit.model <- lm(y4~x4, anscombe)
summary(fit.model)

anscombe$x4_1 <- anscombe$x4 
anscombe$y4_1 <- fit.model$fitted.values -  fit.model$residuals
```

## Original Plots

```{r}
# 2x2 row with 2 margin
par(mfrow=c(2,2), mar=c(2, 2, 2, 2))
plot(anscombe$x1, anscombe$y1, xlim=c(2, 20), ylim=c(2, 14), pch = 19, 
    col = "darkblue", cex = 0.3, las = 1, xlab = "", ylab = "") 
abline(lm(anscombe$y1~anscombe$x1))
title("Set I")

plot(anscombe$x2, anscombe$y2, xlim=c(2, 20), ylim=c(2, 14), pch = 19, 
    col = "darkblue", cex = 0.3, las = 1, xlab = "", ylab = "")
abline(lm(anscombe$y2~anscombe$x2))
title("Set II")

plot(anscombe$x3, anscombe$y3, xlim=c(2, 20), ylim=c(2, 14), pch = 19, 
    col = "darkblue", cex = 0.3, las = 1, xlab = "", ylab = "")
abline(lm(anscombe$y3~anscombe$x3))
title("Set III")

plot(anscombe$x4, anscombe$y4, xlim=c(2, 20), ylim=c(2, 14), pch = 19, 
    col = "darkblue", cex = 0.3, las = 1, xlab = "", ylab = "")
abline(lm(anscombe$y4~anscombe$x4))
title("Set IV")
```

# Modified Anscombe Quartet

To modify the original data while keeping the mean and standard deviation unchaged I fliped the dependent variable on the linear regression line.  The expected impact is that the  $\beta0$ (intercept) and $\beta1$ will not change.  The mean and standard deviation for the explantory variables will not change.

```{r}
kable(anscombe %>% select(-(x1:y4)))
```

## Descriptive statistics

```{r}
# calculate the mean and standard deviation
kable(anscombe %>% select(-(Observation:y4)) %>%
    summarise_all(funs( mean(.), sd(.))) %>%
  gather())
```

## Modified Linear Model

```{r}
fit.model <- lm(y1_1~x1_1, anscombe)
summary(fit.model)

fit.model <- lm(y2_1~x2_1, anscombe)
summary(fit.model)

fit.model <- lm(y3_1~x3_1, anscombe)
summary(fit.model)

fit.model <- lm(y4_1~x4_1, anscombe)
summary(fit.model)
```

## Modified Plots

```{r}
# 2x2 row with 2 margin
par(mfrow=c(2,2), mar=c(2, 2, 2, 2))
plot(anscombe$x1_1, anscombe$y1_1, xlim=c(2, 20), ylim=c(2, 14), pch = 19, 
    col = "darkblue", cex = 0.3, las = 1, xlab = "", ylab = "") 
abline(lm(anscombe$y1_1~anscombe$x1_1))
title("Set I")

plot(anscombe$x2_1, anscombe$y2_1, xlim=c(2, 20), ylim=c(2, 14), pch = 19, 
    col = "darkblue", cex = 0.3, las = 1, xlab = "", ylab = "")
abline(lm(anscombe$y2_1~anscombe$x2_1))
title("Set II")

plot(anscombe$x3_1, anscombe$y3_1, xlim=c(2, 20), ylim=c(2, 14), pch = 19, 
    col = "darkblue", cex = 0.3, las = 1, xlab = "", ylab = "")
abline(lm(anscombe$y3_1~anscombe$x3_1))
title("Set III")

plot(anscombe$x4_1, anscombe$y4_1, xlim=c(2, 20), ylim=c(2, 14), pch = 19, 
    col = "darkblue", cex = 0.3, las = 1, xlab = "", ylab = "")
abline(lm(anscombe$y4_1~anscombe$x4_1))
title("Set IV")
```



