---
title: "Project 5: Operations Management"
author: "Marjorie Blanco"
subtitle: "DA 420"
geometry: "left=1cm,right=3cm,top=2cm,bottom=2cm"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r echo=FALSE}
# Workforce Scheduling for Anonymous Bank Call Center (R)

library(lubridate)  # date functions
library(grid)  # graphics utilities needed for split-plotting
library(ggplot2)  # graphics package with ribbon plot
library(queueing)  # queueing functions, including Erlang C
library(lpSolve)  # linear programming package
library(kableExtra)
library(dplyr)
```

## Problem 1: Short Answers

Part 1: Use R to develop a “Call Center Arrival and Service Rates on Thursday”, similar toFigure 6.7 in page 72. Make sure you include the command lines and the output/results. Interpret the graphin details. 

Hint:Look at the Exhibit 6.1 in page 77


```{r echo=FALSE}
df <- data.frame(Day=character(), 
                 percentage=character(), 
                 stringsAsFactors=FALSE) 
put.title.on.plots <- TRUE  # put title on wait-time ribbon plots
source("Chapter_6/R_utility_program_3.R") #provides split-plotting utilities
source("Chapter_6/R_utility_program_4.R") #provides wait-time ribbon plots

call.center.input.data <- read.table("Chapter_6/data_anonymous_bank_february.txt", 
                                     header = TRUE, colClasses = c("character","integer","numeric",
                                                                   "integer","character","character","character","character","integer",
                                                                   "character","character","integer","factor","character","character",
                                                                   "integer","character"))

# delete PHANTOM calls
call.center.data <- subset(call.center.input.data, subset = (outcome != "PHANTOM"))

# negative VRU times make no sense... drop these rows from data frame
call.center.data <- subset(call.center.data, subset = (vru_time >= 0))

# calculate wait time as sum of vru_time and q_time
call.center.data$wait_time <- call.center.data$vru_time + call.center.data$q_time

# define four-digit year so year is not read as 2099
# convert date string to date variable 
call.center.data$date <- paste("19", call.center.data$date, sep ="")
call.center.data$date <- ymd(call.center.data$date)

# identify day of the week 1 = Sunday ... 7 = Saturday
call.center.data$day_of_week <- wday(call.center.data$date)
call.center.data$day_of_week <- factor(call.center.data$day_of_week,
                                       levels = c(1:7), labels = c("Sunday","Monday","Tuesday",
                                                                   "Wednesday","Thursday","Friday","Saturday"))
```

```{r echo=FALSE}
# identify the hour of entry into the system
time.list <- strsplit(call.center.data$vru_entry,":")
call.hour <- numeric(nrow(call.center.data))
for (index.for.call in 1:nrow(call.center.data)) 
  call.hour[index.for.call] <- as.numeric(time.list[[index.for.call]][1])
call.center.data$call_hour <- call.hour
```

Objective Function:

Minimize total daily personnel cost = 252 * $Shift_1$ + 288 * $Shift_2$ + 180 * $Shift_3$ + 180 * $Shift_4$ + 180 * $Shift_5$ + 288 * $Shift_6$ + 288 * $Shift_7$ + 288 * $Shift_8$

Constraints:

For each hour, the available labor hours must be at least equal to the required labor hours.

Assumptions:

- Bank call center will be closed hours 00 through 05.

- Conversion rate from Israeli shekels to United States dollars was 1 ILS = 3.61 USD in June 2013.

```{r}
index.day <- c("Thursday")
plot.title <- paste(index.day,"Call Center Arrival and Service Rates")

# select every day in February for the queueing model
selected.day <- subset(call.center.data, subset = (day_of_week == index.day))

# compute arrival rate of calls as calls for hour  
# we do not use table() here because some hours could have zero calls
calls.for.hour <- numeric(24)
for(index.for.hour in 1:24) { 
  # 24-hour clock has first hour coded as zero in input data file
  coded.index.for.hour <- index.for.hour - 1  
  this.hour.calls <- 
    subset(selected.day, subset = (call_hour == coded.index.for.hour))  
  if(nrow(this.hour.calls) > 0) 
    calls.for.hour[index.for.hour] <- nrow(this.hour.calls)  
}

# compute arrival rate as average number of calls into VRU per hour
hourly.arrival.rate <- calls.for.hour/4

# service times can vary hour-by-hour due to differences 
# in service requests and individuals calling hour-by-hour
# begin by selecting calls that receive service
selected.day.served <- subset(selected.day, subset = (server != "NO_SERVER"))

hourly.mean.service.time <- numeric(24)
served.for.hour <- numeric(24)
for(index.for.hour in 1:24) { 
  # 24-hour clock has first hour coded as zero in input data file
  coded.index.for.hour <- index.for.hour - 1  
  this.hour.calls <- 
    subset(selected.day.served, subset = (call_hour == coded.index.for.hour))
  if(nrow(this.hour.calls) > 0) {
    served.for.hour[index.for.hour] <- nrow(this.hour.calls)
    hourly.mean.service.time[index.for.hour] <- mean(this.hour.calls$ser_time)
  }
} 

# hourly service rate given the current numbers of service operators
hourly.served.rate <- served.for.hour/4 
# build data frame for plotting arrival and service rates
hour <- 1:24  # hour for horizontal axix of line chart
type <- rep("Arrived", length = 24)
value <- hourly.arrival.rate
arrival.data.frame <- data.frame(hour, value, type) 
type <- rep("Served", length = 24)
value <- hourly.served.rate
service.data.frame <- data.frame(hour, value, type) 
arrival.service.data.frame <- rbind(arrival.data.frame, service.data.frame)

plotting.object <- ggplot(data = arrival.service.data.frame, 
                          aes(x = hour, y = value, fill = type)) + 
  geom_line() +
  geom_point(size = 4, shape = 21) +
  scale_y_continuous(limits = c(0,210)) +
  scale_x_continuous(breaks = c(1,3,5,7,9,11,13,15,17,19,21,23,25),
                     labels = 
                       c("00","02","04","06","08","10","12","14","16","18","20","22","24")) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1, vjust = 1)) +
  labs(x = "Hour of Day (24-Hour Clock)", y = "Average Calls per Hour") +
  labs(title = plot.title) +
  scale_fill_manual(values = c("yellow","dark green"), 
                    guide = guide_legend(title = NULL))  +
  theme(legend.position = c(1,1), legend.justification = c(1,1)) +
  theme(legend.text = element_text(size=15)) +
  coord_fixed(ratio = 1/10)
print(plotting.object)

plot.title <- paste(index.day, "Call Center Needs and Optimal Workforce Schedule")

# select every day in February for the queueing model
# selected.day <- subset(call.center.data, subset = (day_of_week == this.day.of.week))
# 
# selected.day.served <- subset(selected.day, subset = (server != "NO_SERVER"))

# examine service times per service operator
# for hours with no service time information use the mean as value
hourly.mean.service.time <- 
  ifelse((hourly.mean.service.time == 0),
         mean(selected.day.served$ser_time),
         hourly.mean.service.time) 
# compute service rate noting that there are 3600 seconds in an hour
# adding 60 seconds to each mean service time for time between calls
# this 60 seconds is the wrap up time or time a service agent remains 
# unavailable to answer a new call after a call has been completed
hourly.service.rate <- 3600/(hourly.mean.service.time + 60)

# we observe that mean service times do not vary that much hour-by-hour
# so we use the mean hourly service rate in queueing calculations

cat("\n","Hourly service rate:")
print(mean(hourly.service.rate))
SERVICE.RATE <- round(mean(hourly.service.rate), 0)

# C_erlang function from the queueing package
# inputs c = number of servers
#        r = ratio of rate of arrivals and rate of service
# returns the propability of waiting in queue because all servers are busy
# let us set a target for the probability of waiting in queue to be 0.50
# using while-loop iteration we determine the number of servers needed 
# we do this for each hour of the day knowing the hourly arrival rate

PROBABILITY.GOAL <- 0.50
servers.needed <- integer(24)  # initialize to zero
for(index.for.hour in 1:24) {
  if (hourly.arrival.rate[index.for.hour] > 0) {
    erlang.probability <- 1.00  # intialization prior to entering while-loop
    while (erlang.probability > PROBABILITY.GOAL) {
      servers.needed[index.for.hour] <- servers.needed[index.for.hour] + 1
      erlang.probability <- C_erlang(c = servers.needed[index.for.hour], 
                                     r = hourly.arrival.rate[index.for.hour]/SERVICE.RATE)
    }  # end while-loop for defining servers needed given probability goal 
  }  # end if-block for hours with calls
}  # end for-loop for the hour

# the result for servers.needed is obtained as
# 1  1  1  0  1  1  1  4  8  9 10  9  8 16 10 10  6  7  8  8  6  6  5  4
# we will assume the bank call center will be closed hours 00 through 05
# but use the other values as the bank's needed numbers of servers
servers.needed[1:6] <- 0

cat("\n",index.day, "Hourly Operator Requirements:","\n")
print(servers.needed)

# read in case data for the structure of call center worker shifts
bank.shifts.data.frame <- read.csv("Chapter_6/data_anonymous_bank_shifts.csv")


constraint.matrix <- as.matrix(bank.shifts.data.frame[,3:10])

constraint.matrix1 <- cbind(constraint.matrix, Need = servers.needed)
cat("\n",index.day, "Call Center Shift Constraint Matrix","\n")
print(constraint.matrix1)

# six-hour shift salaries in Israeli sheqels 
# 1 ILS = 3.61 USD in June 2013
# these go into the objective function for integer programing
# with the objective of minimizing total costs
cost.vector <- c(252,288,180,180,180,288,288,288) 

call.center.schedule <- lp(const.mat=constraint.matrix,
                           const.rhs = servers.needed,
                           const.dir = rep(">=",times=8),
                           int.vec = 1:8,
                           objective = cost.vector,
                           direction = "min")

# prepare summary of the results for the call center problem
ShiftID <- 1:8
StartTime <- c(0,6,8,10,12,2,4,6)
# c("Midnight","6 AM","8 AM","10 AM","Noon","2 PM","4 PM","6 PM")
ShiftDuration <- rep(6,times=8)
HourlyShiftSalary <- c(42,48,30,30,30,48,48,48)
HourlyShiftCost <- call.center.schedule$objective # six x hourly shift salary
Solution <- call.center.schedule$solution  
ShiftCost <- call.center.schedule$solution * call.center.schedule$objective

call.center.summary <- 
  data.frame(ShiftID,StartTime,ShiftDuration,HourlyShiftSalary,
             HourlyShiftCost,Solution,ShiftCost)

cat("\n\n",index.day,"Call Center Summary","\n\n")
print(call.center.summary)

# the solution is obtained by print(call.center.schedule) 
# or by summing across the hourly solution times the cost objective'
#cat("\n\n","Call Center Schedule","\n\n")
#print(call.center.schedule) 
cat("\n\n",index.day,"Call Center Summary Minimum Cost Solution:",sum(ShiftCost),"\n\n")
# build data frame for plotting the solution compared with need
hour <- 1:24  # hour for horizontal axix of line chart
type <- rep("Hourly Need", length = 24)
value <- servers.needed
needs.data.frame <- data.frame(hour, value, type) 
type <- rep("Optimal Solution", length = 24)
value <- schedule.fit.to.need <- 
  constraint.matrix %*% call.center.schedule$solution
solution.data.frame <- data.frame(hour, value, type) 
plotting.data.frame <- rbind(needs.data.frame, solution.data.frame)

# plot the solution... solution match to the workforce need
plotting.object <- ggplot(data = plotting.data.frame, 
                          aes(x = hour, y = value, fill = type)) + 
  geom_line() +
  geom_point(size = 4, shape = 21) +
  scale_y_continuous(limits = c(0,20)) +
  scale_x_continuous(breaks = c(1,3,5,7,9,11,13,15,17,19,21,23,25),
                     labels = 
                       c("00","02","04","06","08","10","12","14","16","18","20","22","24")) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1, vjust = 1)) +
  labs(x = "Hour of Day (24-Hour Clock)", y = "Number of Service Operators") +
  labs(title = plot.title) +
  scale_fill_manual(values = c("white","blue"), 
                    guide = guide_legend(title = NULL)) +
  theme(legend.position = c(1,1), legend.justification = c(1,1)) +
  theme(legend.text = element_text(size=15)) +
  coord_fixed(ratio = 2/2.25)    
print(plotting.object)
```

